---
title: "MAC Microbiome Study"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE)

# Packages used
pacs <- c("tidyverse", "readxl", "haven", "tableone", "gtsummary", "gt", "ggbeeswarm", "tidytext", "kableExtra")
sapply(pacs, require, character.only = TRUE)

```

```{r read_data, include=FALSE}

# Read previous MAC data --------------------------------------------------

# Needed to extract treatment sequence
mac0 <- read_spss("./data/MAC Endpoint data with baseline values 102121.sav") %>% 
  janitor::clean_names()

# Variables needed
seq <- mac0 %>% 
  select(id, group, phase, visit, treatment, gender, age_b, weight_b)

# Distinct data with ID, Group, gender, age, weight
# n = 35
seq_distinct <- seq %>% 
  select(id, group, gender, age_b, weight_b) %>% 
  distinct()

# Read cardiometabolic data -----------------------------------------------

# Variable list
cm_vars <- c(
  "glucose",
  "insulin",
  "chol",
  "trigs",
  "hdl",
  "ldl",
  "vldl",
  "apoa1",
  "apob",
  "crp",
  "eselectin"
)

# 108 obs x 15 variables
cm <- read_excel("./data/LLU_MAC Study_cardiometabolic outcomes.xlsx", sheet = "Data copy") %>% 
  setNames(c("id", "phase", "visit", "clinic_date", cm_vars)) %>%  
  mutate(clinic_date = as.Date(clinic_date))

# Read anthropometric data ------------------------------------------------

# Needed for adiposity variables
# 35 subj x 23 vars
adip <- read_excel("./data/MAC Anthropometric data.xlsx") %>% 
  janitor::clean_names() %>% 
  rename(id = study_id)

# Make it to long format
# Should have 35 * 3 = 105 obs
adip_long <- adip %>% 
  select(
    id, 
    visit_0_in_body_bmi, 
    clinic_4_in_body_bmi,
    clinic_9_wk_18_in_body_bmi,
    visit_0_in_body_percent_body_fat,
    clinic_4_in_body_percent_body_fat,
    clinic_9_wk_18_in_body_percent_body_fat
  ) %>%
  rename(
    bmi_0     = visit_0_in_body_bmi,
    bmi_4     = clinic_4_in_body_bmi,
    bmi_9     = clinic_9_wk_18_in_body_bmi,
    pct_fat_0 = visit_0_in_body_percent_body_fat,
    pct_fat_4 = clinic_4_in_body_percent_body_fat,
    pct_fat_9 = clinic_9_wk_18_in_body_percent_body_fat
  ) %>%
  pivot_longer(
    cols = -id,
    names_to = c(".value", "visit"),
    names_pattern = "(bmi|pct_fat)_(\\d+)"
  ) %>%
  mutate(visit = as.numeric(visit)) %>%
  arrange(id, visit)


# Read microbiome data ----------------------------------------------------

# 4166 rows x 112 columns
mb <- read_excel("./data/human_mac_study_abundance_data.xls")  

# Select "Roseburia" and make it to long format
# Results in 105 obs x 47 vars
roseburia_long <- mb %>%
  filter(str_starts(Species, "Roseburia")) %>%
  select(Species, starts_with("MAC")) %>%
  pivot_longer(
    cols = -Species,
    names_to = "sample",
    values_to = "abundance"
  ) %>%
  mutate(
    id    = as.integer(str_extract(sample, "(?<=MAC)\\d+")),
    visit = as.integer(str_extract(sample, "(?<=v)\\d+"))
  ) %>%
  select(-sample) %>%
  pivot_wider(
    id_cols    = c(id, visit),
    names_from = Species,
    values_from = abundance,
    values_fn  = sum   # combines rows if >1 OTU/ASV shares the same Species label
  ) %>%
  arrange(id, visit) %>% 
  janitor::clean_names()


# Merge data --------------------------------------------------------------

df <- adip_long %>% 
  inner_join(cm, by = c("id", "visit")) %>% 
  left_join(seq_distinct, by = "id") %>% 
  inner_join(roseburia_long, by = c("id", "visit")) %>% 
  mutate(
    group_lab = factor(group, labels = c("mac-control", "control-mac")),
    phase     = ifelse(visit == 0, 0, phase),
    treatment = case_when(
      visit == 0                                            ~ "baseline",
      (group == 1 & phase == 1) | (group == 2 & phase == 2) ~ "mac",
      (group == 1 & phase == 2) | (group == 2 & phase == 1) ~ "control"
    )
  ) %>% 
  select(
    id, 
    group, 
    group_lab, 
    clinic_date, 
    visit, 
    phase, 
    treatment,
    gender,
    age_b,
    weight_b,
    bmi, 
    pct_fat,
    chol,
    ldl, 
    hdl,
    apob,
    starts_with("roseburia")
  )

```

## Datasets 

* Data are from a 2 x 2 crossover design (AB/BA) with control/mac treatments ([Jones et al., 2023](https://doi.org/10.1017/jns.2023.39))
 
* Anthropometric data
  * Excel file `MAC Anthropometric data.xlsx`
  * Wide-format data for 35 unique participant IDs
  * Includes **BMI** and **percent body fat** (at `visit 0`, `clinic 4`, and `clinic 9`)
    * Also includes: weight (kg) and waist circumference (cm)
 
* Cardiometabolic data
  * Excel file `LLU_MAC Study_cardiometabolic outcomes.xlsx`
  * Long-format data for 38 unique participant IDs
    * Baseline (visit 0), Phase 1 (visit 4), Phase 2 (visit 9), and clinic date
  * Includes **total cholesterol**, **LDL**, **HDL**, and **ApoB**
    * Also includes: glucose, insulin, triglycerides, VLDL, ApoA1, CRP, and E-selectin
 
* Microbiome data
  * Excel file `human_mac_study_abundance_data.xls`
  * Wide-format data with 4,166 bacteria species/ASVs, including **Roseburia**
  * The first 7 columns are: `Domain`, `Phylum`, `Class`, `Order`, `Family`, `Genus`, and `Species`
  * Abundance values at visits 0, 4, and 9 are stored in the 105 columns that follow, named:
    * MAC[_ID_in_2digits_]_p1[_v0/v4/v9_]
    
* To extract participants' treatment sequence, use:
  * SPSS data file `MAC Endpoint data with baseline values 102121.sav`
  * Participant characteristics were also extracted from this file
    * Sex, age at baseline, and weight at baseline
    
* Data files were merged to produce a long-format dataset
  * 105 observations (3 visits, including baseline, × 35 participants)


## Descriptive table at baseline

* Descriptive statistics at baseline by sequence
  * [**Need to check**] Descriptive statistics on percent body fat are slightly off, compared to the original paper ([Jones et al., 2023](https://doi.org/10.1017/jns.2023.39))

```{r descriptive_table}

# Using gtsummary
pval_4dp <- function(x) {
  ifelse(x < 0.0001, "<0.0001", sprintf("%.4f", x))
}

tbl <- df %>% 
  filter(visit == 0) %>% 
  select(group_lab, age_b, gender, weight_b, bmi, pct_fat, chol, ldl, hdl, apob) %>% 
  tbl_summary(
    by = group_lab,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 1,
    label  = list(
      age_b    ~ "Age (years)",
      gender   ~ "Sex",
      weight_b ~ "Weight (kg)",
      bmi      ~ "BMI (kg/m²)",
      pct_fat  ~ "Body fat (%)",
      chol     ~ "Total cholesterol (mg/dL)",
      ldl      ~ "LDL cholesterol (mg/dL)",
      hdl      ~ "HDL cholesterol (mg/dL)",
      apob     ~ "ApoB (mg/dL)"
    )
  ) %>% 
  add_overall() %>% 
  add_p(
    test = list(
      all_continuous()  ~ "t.test",
      all_categorical() ~ "fisher.test"
    ),
    pvalue_fun = pval_4dp
  ) %>% 
  bold_labels() %>% 
  modify_footnote(everything() ~ NA) %>% 
  as_kable_extra() %>% 
  footnote(
    general = "Values are mean (SD) for continuous variables and n (%) for categorical variables. P-values from Welch's two-sample t-test (continuous) or Fisher's exact test (categorical).",
    general_title = "",
    footnote_as_chunk = TRUE
  )

tbl

```

## Exploratory analyses

```{r df_delta}

# Time-invariant covariates + baseline BMI/%fat, from the visit == 0 row
baseline_df <- df %>% 
  filter(visit == 0) %>% 
  select(id, group, group_lab, gender, age_b, weight_b, 
         bmi_b = bmi, pct_fat_b = pct_fat)

# Mac vs. Control values, reshaped wide and differenced
delta_df <- df %>% 
  filter(treatment %in% c("mac", "control")) %>% 
  select(id, treatment, chol, ldl, hdl, apob, roseburia_2) %>% 
  pivot_wider(
    id_cols     = id,
    names_from  = treatment,
    values_from = c(chol, ldl, hdl, apob, roseburia_2)
  ) %>% 
  mutate(
    delta_chol        = chol_mac        - chol_control,
    delta_ldl         = ldl_mac         - ldl_control,
    delta_hdl         = hdl_mac         - hdl_control,
    delta_apob        = apob_mac        - apob_control,
    delta_roseburia_2 = roseburia_2_mac - roseburia_2_control
  ) %>% 
  select(id, starts_with("delta_"))

# Combine into the final one-row-per-participant analytic dataset
df_delta <- baseline_df %>% 
  inner_join(delta_df, by = "id")

```

### Cardiometabolic variables

* Distributions of cardiometabolic measurements (excluding baseline)

```{r histogram_cardiometabolic}

# Histogram
df %>%
  filter(phase != 0) %>% 
  pivot_longer(cols = c("ldl", "chol", "hdl", "apob"), names_to = "var", values_to = "value") %>% 
  mutate(
    var = factor(var, 
    levels = c("chol", "ldl", "hdl", "apob"), 
    labels = c("Total cholesterol (mg/dL)", "LDL cholelsterol (mg/dL)", "HDL cholesterol (mg/dL)", "ApoB (mg/dL)")) 
  ) %>% 
  ggplot(aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20) +
  geom_density(color = "cornflowerblue", linewidth = 1, adjust = 2) +
  labs(x = NULL, y = "Density") +
  facet_wrap(~ var, scales = "free", ncol = 2) +
  theme_bw()

```

* Profile plots of cardiometabolic measurements by sequence
  * Each line represents a participant
  * Visualize individual-level change across phases
    * All within-person changes across phases fall within a physiologically plausible range

```{r profile_plots_cardiometabolic, fig.width = 6, fig.height = 10}

# Profile plot
df %>% 
  filter(visit != 0) %>%  
  mutate(phase = factor(phase)) %>% 
  pivot_longer(cols = c("ldl", "chol", "hdl", "apob"), names_to = "var", values_to = "value") %>% 
  mutate(
    var = factor(var, 
    levels = c("chol", "ldl", "hdl", "apob"), 
    labels = c("Total cholesterol (mg/dL)", "LDL cholelsterol (mg/dL)", "HDL cholesterol (mg/dL)", "ApoB (mg/dL)")) 
  ) %>% 
  ggplot(aes(x = phase, y = value, group = id, color = group_lab)) +
  geom_point() +
  geom_line() +
  scale_x_discrete(breaks = 1:2, expand = expansion(add = 0.2)) +
  labs(x = "Phase", y = NULL) +
  facet_grid(var ~ group_lab, scales = "free", switch = "y", axes = "all_x") +
  theme_bw() +
  theme(
    legend.position = "none",
    strip.placement = "outside",
    strip.background = element_blank()
  ) 

```

* Waterfall plots of within-person change (mac − control) across four cardiometabolic outcomes
  * Each bar represents one participant's change score, sorted from most negative to most positive
  * Visualize the direction and magnitude of individual treatment response, and whether response is consistent across the cohort or driven by a subset of participants

```{r waterfall_plot_cardiometabolic, fig.width = 12, fig.height = 7}

# Barplot on delta
df_delta %>% 
  pivot_longer(
    cols = c(delta_chol, delta_ldl, delta_hdl, delta_apob), 
    names_to = "var", 
    values_to = "value"
  ) %>% 
  mutate(
    var = factor(var, 
                 levels = c("delta_chol", "delta_ldl", "delta_hdl", "delta_apob"),
                 labels = c("Δ Total cholesterol (mg/dL)", "Δ LDL (mg/dL)", 
                            "Δ HDL (mg/dL)", "Δ ApoB (mg/dL)")),
    id_ordered = reorder_within(id, value, var)
  ) %>% 
  ggplot(aes(x = id_ordered, y = value, fill = group_lab)) +
  geom_bar(stat = "identity") +
  scale_x_reordered() +
  facet_wrap(~ var, scales = "free") +
  labs(
    x = "Participant ID",
    y = "Difference: Macademia - Control", 
    fill = "Treatement sequence" 
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

```

* Mean (SD) within-person change (mac − control) for total cholesterol, LDL, HDL, and ApoB

```{r mean_delta_cardiometabolic}


# Build mac/control/delta values in one wide frame
delta_full <- df %>% 
  filter(treatment %in% c("mac", "control")) %>% 
  select(id, treatment, chol, ldl, hdl, apob) %>% 
  pivot_wider(
    id_cols     = id,
    names_from  = treatment,
    values_from = c(chol, ldl, hdl, apob)
  ) %>% 
  mutate(
    chol_delta = chol_mac - chol_control,
    ldl_delta  = ldl_mac  - ldl_control,
    hdl_delta  = hdl_mac  - hdl_control,
    apob_delta = apob_mac - apob_control
  )

# Reshape to long (outcome x condition), summarize, then pivot back to wide
summary_tbl <- delta_full %>% 
  pivot_longer(
    cols         = -id,
    names_to     = c("outcome", "condition"),
    names_pattern = "(chol|ldl|hdl|apob)_(mac|control|delta)",
    values_to    = "value"
  ) %>% 
  group_by(outcome, condition) %>% 
  summarise(mean_sd = sprintf("%.1f (%.1f)", mean(value), sd(value)), .groups = "drop") %>% 
  pivot_wider(names_from = condition, values_from = mean_sd) %>% 
  mutate(
    outcome = factor(outcome, 
                     levels = c("chol", "ldl", "hdl", "apob"),
                     labels = c("Total cholesterol (mg/dL)", "LDL cholesterol (mg/dL)", 
                                "HDL cholesterol (mg/dL)", "ApoB (mg/dL)"))
  ) %>% 
  arrange(outcome) %>% 
  select(Outcome = outcome, Mac = mac, Control = control, `Δ (Mac-Control)` = delta)

summary_tbl %>% knitr::kable(align = c("l", "c", "c", "c"))

```

### Microbiome variables

* Only focusing on **Roseburia_2** for now. See the distribution below, excluding baseline:
  * The x-axis is on the pseudo-log scale
  * Note that there are >25 zero counts

```{r histogram_roseburia_2, fig.height = 4, fig.width = 5}

df %>%
  filter(phase != 0) %>% 
  ggplot(aes(x = roseburia_2)) +
  geom_histogram() +
  scale_x_continuous(transform = scales::pseudo_log_trans(), breaks = c(0, 10, 30, 100, 300, 1000, 3000)) +
  labs(x = "Roseburia_2 (pseudo-log scale)") +
  theme_bw()
 
```

* There are 11 participants with **Roseburia_2** = 0 in both phases
* Six participants have **Roseburia_2 = 0** in only one of the two phases
  * Incidentally, in all six cases, the zero occurred under the control treatment, not the macadamia treatment

```{r check_zero_roseburia_2}

# Checking IDs with Roseburia_2 = 0
zero_check <- df %>% 
  filter(phase %in% c(1, 2)) %>% 
  select(id, phase, group_lab, roseburia_2) %>% 
  pivot_wider(names_from = phase, values_from = roseburia_2, names_prefix = "phase_")

both_zero_tbl <- zero_check %>% 
  filter(phase_1 == 0 & phase_2 == 0) %>% 
  select(id, group_lab, phase_1, phase_2) %>% 
  kable(caption = "Zero in both phases")

one_zero_tbl <- zero_check %>% 
  filter(xor(phase_1 == 0, phase_2 == 0)) %>% 
  select(id, group_lab, phase_1, phase_2) %>% 
  kable(caption = "Zero in one phase")

knitr::kables(list(both_zero_tbl, one_zero_tbl))

```

* Profile plots of **Roseburia_2** by sequence
  * Shows higher **Roseburia_2** abundance under the macadamia diet
  * The flat lines at zero correspond to the 11 participants with Roseburia_2 = 0 in both phases (see table above)
  * Note: the y-axis uses a pseudo-log scale to accommodate true zeros, so visual distances near the bottom of the axis are compressed/expanded non-linearly relative to distances higher up
    * A line rising from 0 shouldn't be read as directly comparable in magnitude to, say, a line rising from 300 to 1000

```{r profile_plot_roseburia_2, fig.height = 4}
# Profile plot
df %>% 
  filter(phase != 0) %>%
  mutate(phase = factor(phase)) %>% 
  ggplot(aes(x = phase, y = roseburia_2, group = id, color = group_lab)) +
  geom_point() +
  geom_line() +
  scale_x_discrete(breaks = 1:2, expand = expansion(add = 0.1)) +
  scale_y_continuous(transform = scales::pseudo_log_trans(), breaks = c(0, 10, 30, 100, 300, 1000, 3000)) +
  labs(
    x     = "Phase", 
    y     = "Roseburia_2 (pseudo-log scale)", 
    color = "Treatment sequence"
  ) +
  facet_wrap(~ group_lab) +
  theme_bw() +
  theme(legend.position = "bottom")

```

* Median (IQR) of **Roseburia_2** by treatment, presented here (rather than mean/SD or geometric mean) given the substantial zero-inflation in this variable (see zero-count table above)

```{r descriptives_roseburia_2}

# Descriptive stats
df %>% 
  filter(treatment %in% c("mac", "control")) %>% 
  select(id, treatment, roseburia_2) %>% 
  pivot_wider(id_cols = id, names_from = treatment, values_from = roseburia_2) %>% 
  mutate(roseburia_2_delta = mac - control) %>% 
  pivot_longer(-id, names_to = "treatment", values_to = "value") %>% 
  mutate(treatment = factor(treatment, levels = c("mac", "control", "roseburia_2_delta"))) %>% 
  group_by(treatment) %>% 
  summarise(median_iqr = sprintf("%.1f (%.1f)", median(value), IQR(value)), .groups = "drop") %>% 
  pivot_wider(names_from = "treatment", values_from = median_iqr) %>% 
  rename(`Mac, Median (IQR)` = mac, `Control, Median (IQR)` = control, `Δ Roseburia_2 (Mac-Control)` = roseburia_2_delta) %>% 
  knitr::kable()

```

### Bivariate relationship with cardiometabolic measures

* To examine the association between Δ **Roseburia_2** and Δ total cholesterol, LDL, HDL, and ApoB, scatterplots were generated for each pair
  * Participants were split into two groups by the median baseline % body fat (<43% and ≥43%), and a linear trend line was overlaid for each group
* In the scatterplots for Δ total cholesterol, LDL, and ApoB, slopes are visibly different between the two adiposity groups: a stronger negative association appears in the low adiposity group (<43%), while the high adiposity group (≥43%) shows a slope close to flat
  * HDL shows little relationship with Δ Roseburia_2 in either group
* These patterns suggest that baseline %body fat may modify the association between Δ Roseburia_2 and these cardiometabolic outcomes

```{r scatterplot_with_cardiometabolics, fig.width = 8, fig.height = 6}

# Scatterplot of deltas
df_delta %>% 
  select(id, group, pct_fat_b, starts_with("delta_")) %>% 
  mutate(
    pct_fat_b_cat = ifelse(pct_fat_b < 43, 0, 1),
    pct_fat_b_cat = factor(pct_fat_b_cat, labels = c("<43%", "≥43%"))
  ) %>% 
  pivot_longer(
    cols = c(delta_chol, delta_ldl, delta_hdl, delta_apob),
    names_to = "var",
    values_to = "value"
  ) %>% 
  mutate(
    var = factor(var, 
                 levels = c("delta_chol", "delta_ldl", "delta_hdl", "delta_apob"),
                 labels = c("Total cholesterol", "LDL", "HDL", "ApoB"))
  ) %>% 
  ggplot(aes(x = delta_roseburia_2, y = value, color = pct_fat_b_cat)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ var, scales = "free_y") +
  labs(
    x = "Δ Roseburia_2 (Mac − Control)", 
    y = "Δ outcome (Mac − Control)",
    color = "% fat at baselinie"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

```


