---
title: "Power simulation and misc analyses for grant"
output: github_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE)

# Packages used

pacs <- c(
  "jsonlite",
  "tidyverse",
  "readxl", 
  "haven",
  "irr",
  "vegan",
  "ggbeeswarm"
)

sapply(pacs, require, character.only = TRUE)

```


## Power simulation

* To project statistical power for the primary analysis (the ETI × capacity score interaction) at the anticipated evaluable sample size, we conducted a Monte Carlo power simulation with 20,000 replicates per scenario
  * Each replicate simulated a dataset under the assumed data-generating model:
    * A reduced model (ETI, capacity score, and **6 covariates**) explaining R² = 0.20 of outcome variance, 
    * The ETI x capacity interaction term targeting an incremental effect size of f² = 0.029 (approximately ΔR² = 0.023)
    * Power was defined as the proportion of replicates in which the interaction term was statistically significant using HC3 robust standard errors

* Power was estimated under three capacity-sampling modes
  * Resampling directly from the observed empirical distribution (900 values)
  * A winsorized version (1st/99th percentile trimmed) to check sensitivity to potential outliers
  * A normal approximation matched to the empirical mean and SD
  
* The table below reflects the **uncorrelated** scenario, with ETI and capacity score treated as statistically independent
  * ETI-SR was simulated as an independent standard normal ($ETI_i \sim \mathcal{N}(0,1)$) in this uncorrelated scenario, consistent with the assumption of no association between dietary/behavioral exposure and capacity score
  * At the projected evaluable sample size of N = 328, power exceeded 80% across all three capacity-sampling modes
  * At the accrual-shortfall contingencies (N=300 and N=270), power falls to 77.0–79.2% and 72.2–74.4% respectively

```{r power_primary_uncorrelated}

# Read json file containing simulation results
power_results <- fromJSON("full_run_v3_output.json")

power_results$uncorrelated_grid %>%
  transmute(
    N                        = n,
    Mode                     = mode,
    `Target f²`              = f2,
    Power                    = sprintf("%.1f%%", power_hc3 * 100),
    MCSE                     = sprintf("%.2f pp", mcse * 100),
    # `Realized R² (reduced)`  = round(mean_realized_r2_reduced, 3),
    # `Realized f²`            = round(mean_realized_f2, 3)
  ) %>%
  knitr::kable(align = c("c", "l", "c", "c", "c"))

```

* To evaluate robustness to a plausible **correlation structure **among predictors, we repeated the simulation using a Gaussian copula to induce dependence between ETI-SR, capacity score, and age, while preserving each variable's observed marginal distribution
  * Both ETI-SR and capacity score were resampled from their observed empirical distributions via the copula's quantile transform: $X_i = \hat{Q}_X(u_i)$, where $u_i$ is the copula-transformed uniform variate and $\hat{Q}_X$ is the empirical quantile function for ETI-SR (n = 257) or capacity score (n = 900)
  * The correlation between capacity score and age ($\rho = 0.043$) was fixed at its observed (anchored) value throughout
  * The correlation between ETI-SR and capacity score was varied across $\rho \in \{-0.15, 0, 0.15\}$, and between capacity score and fiber intake across $\rho \in \{0, 0.20, 0.40\}$, yielding a 3×3 grid of correlation scenarios
  * This grid was evaluated at all three sample sizes (N = 328, 300, 270), for 27 scenarios total, each with 20,000 replicates

* Power was largely insensitive to the capacity–fiber correlation (≤1 pp shift across ρ = 0–0.40 at any fixed N and ETI–capacity correlation)
  * Power was more sensitive to the direction of the ETI–capacity correlation: a negative correlation (ρ = −0.15) increased power by roughly 3 pp relative to independence, while a positive correlation (ρ = 0.15) decreased power by roughly 3–4 pp, at every sample size evaluated
  * Across the full correlated grid, power at the primary sample size (N = 328) ranged from 77.3% to 84.1%. At the accrual-shortfall sample sizes, the lowest observed power was 67.9-68.3% (N = 270, positive ETI–capacity correlation)

```{r power_correlated_N328}

power_results$correlated_grid %>%
  transmute(
    N                    = n,
    `ρ (ETI–capacity)`   = rho_eti_capacity,
    `ρ (capacity–fiber)` = rho_capacity_fiber,
    Power                = sprintf("%.1f%%", power_hc3 * 100),
    MCSE                 = sprintf("%.2f pp", mcse * 100)
  ) %>%
  knitr::kable(align = "c")

```

```{r power_correlated_N300}

```

## MAC 16S secondary analyses

### Intra-class correlation on Shannon Index

* A two-way, absolute-agreement, single-measure intraclass correlation coefficient (ICC) was calculated to assess test-retest reliability of the Shannon diversity index, comparing values measured at baseline and at the end of the control phase among participants in the control-mac sequence group

* The ICC was 0.954 (95% CI: 0.878–0.983; p < 0.0001), indicating excellent test-retest reliability
  * This suggests that overall gut microbial community diversity, as summarized by the Shannon index, remained highly stable within individuals over the course of the control phase, supporting its use as a reproducible, individual-level measure of diversity in this cohort.

```{r ICC_prep, include=FALSE}

# Read previous MAC data --------------------------------------------------

# Needed to extract treatment sequence
mac0 <- read_spss("../data/MAC Endpoint data with baseline values 102121.sav") %>% 
  janitor::clean_names()

# Variables needed
seq <- mac0 %>% 
  select(id, group, phase, visit, treatment, gender, age_b, weight_b)

# Distinct data with ID, Group, gender, age, weight
# n = 35
seq_distinct <- seq %>% 
  select(id, group, gender, age_b, weight_b) %>% 
  distinct()

# Read cardiometabolic data -----------------------------------------------

# Variable list
cm_vars <- c(
  "glucose",
  "insulin",
  "chol",
  "trigs",
  "hdl",
  "ldl",
  "vldl",
  "apoa1",
  "apob",
  "crp",
  "eselectin"
)

# 108 obs x 15 variables
cm <- read_excel("../data/LLU_MAC Study_cardiometabolic outcomes.xlsx", sheet = "Data copy") %>% 
  setNames(c("id", "phase", "visit", "clinic_date", cm_vars)) %>%  
  mutate(clinic_date = as.Date(clinic_date))

# Check distinct IDs: 38 participants
n_distinct(cm$id)

# Read anthropometric data ------------------------------------------------

# Needed for adiposity variables
# 35 subj x 23 vars
adip <- read_excel("../data/MAC Anthropometric data.xlsx") %>% 
  janitor::clean_names() %>% 
  rename(id = study_id)

# Make it to long format
# Should have 35 * 3 = 105 obs
adip_long <- adip %>% 
  select(
    id, 
    visit_0_in_body_bmi, 
    clinic_4_in_body_bmi,
    clinic_9_wk_18_in_body_bmi,
    visit_0_in_body_percent_body_fat,
    clinic_4_in_body_percent_body_fat,
    clinic_9_wk_18_in_body_percent_body_fat
  ) %>%
  rename(
    bmi_0     = visit_0_in_body_bmi,
    bmi_4     = clinic_4_in_body_bmi,
    bmi_9     = clinic_9_wk_18_in_body_bmi,
    pct_fat_0 = visit_0_in_body_percent_body_fat,
    pct_fat_4 = clinic_4_in_body_percent_body_fat,
    pct_fat_9 = clinic_9_wk_18_in_body_percent_body_fat
  ) %>%
  pivot_longer(
    cols = -id,
    names_to = c(".value", "visit"),
    names_pattern = "(bmi|pct_fat)_(\\d+)"
  ) %>%
  mutate(visit = as.numeric(visit)) %>%
  arrange(id, visit)

# Read microbiome data ----------------------------------------------------

# 4166 rows x 112 columns
mb <- read_excel("../data/human_mac_study_abundance_data.xls")  

# Select "Roseburia" and make it to long format
# Results in 105 obs x 47 vars
roseburia_long <- mb %>%
  filter(str_starts(Species, "Roseburia")) %>%
  select(Species, starts_with("MAC")) %>%
  pivot_longer(
    cols = -Species,
    names_to = "sample",
    values_to = "abundance"
  ) %>%
  mutate(
    id    = as.integer(str_extract(sample, "(?<=MAC)\\d+")),
    visit = as.integer(str_extract(sample, "(?<=v)\\d+"))
  ) %>%
  select(-sample) %>%
  pivot_wider(
    id_cols    = c(id, visit),
    names_from = Species,
    values_from = abundance,
    values_fn  = sum 
  ) %>%
  arrange(id, visit) %>% 
  janitor::clean_names()

# Read Shannon index data -------------------------------------------------

# 105 rows x 12 columns
shannon <- read_csv("../data/metadata_with_alpha_div_indices.csv") %>% 
  janitor::clean_names() %>% 
  mutate(
    id    = parse_number(participant_id),
    visit = parse_number(visit)
  ) %>% 
  select(id, visit, shannon)

# Merge data --------------------------------------------------------------

# Group = 1: mac-control
# Group = 2: control-mac

df <- adip_long %>% 
  inner_join(cm,             by = c("id", "visit")) %>% 
  left_join(seq_distinct,    by = c("id"))          %>% 
  inner_join(roseburia_long, by = c("id", "visit")) %>% 
  inner_join(shannon,        by = c("id", "visit")) %>% 
  mutate(
    group_lab = factor(group, labels = c("mac-control", "control-mac")),
    phase     = ifelse(visit == 0, 0, phase),
    treatment = case_when(
      visit == 0                                            ~ "baseline",
      (group == 1 & phase == 1) | (group == 2 & phase == 2) ~ "mac",
      (group == 1 & phase == 2) | (group == 2 & phase == 1) ~ "control"
    )
  ) %>% 
  select(
    id, 
    group, 
    group_lab, 
    clinic_date, 
    visit, 
    phase, 
    treatment,
    gender,
    age_b,
    weight_b,
    bmi, 
    pct_fat,
    chol,
    ldl, 
    hdl,
    apob,
    shannon,
    starts_with("roseburia")
  )

```


```{r ICC}

# ICC ---------------------------------------------------------------------

# ICC(2, 1) for absolute agreement
# Use only control-first group
# Compare at baseline and at the end of control phase
df_icc <- df %>% 
  select(id, group, group_lab, visit, phase, shannon) %>% 
  filter(group == 2, phase %in% c(0, 1))

# Make it to wide format
df_icc_wide <- df_icc %>% 
  select(id, phase, shannon) %>% 
  pivot_wider(names_from = phase, values_from = shannon, names_prefix = "phase_")

# Two-way, single-measure ICC for absolute agreement
icc_result <- df_icc_wide %>% 
  select(-id) %>% 
  icc(model = "twoway", type = "agreement")

icc_table <- tibble(
  Subjects   = icc_result$subjects,
  Raters     = icc_result$raters,
  `ICC(A,1)` = round(icc_result$value, 3),
  `95% CI`   = sprintf("%.3f, %.3f", icc_result$lbound, icc_result$ubound),
  `F`        = round(icc_result$Fvalue, 1),
  df1        = icc_result$df1,
  df2        = icc_result$df2,
  `p-value`  = ifelse(icc_result$p.value < 0.001, "< 0.001",
                       sprintf("%.3g", icc_result$p.value))
)

knitr::kable(
  icc_table,
  align   = "c"
)

```

### Bray-Curtis similarity, within- vs between-person

* To evaluate compositional reliability, we calculated Bray-Curtis similarity between all pairs of stool samples collected from participants in the control-first (control-mac) sequence group, based on relative abundance across 4,166 taxa
  * Similarity values range from 0 (no shared taxa) to 1 (identical composition), with higher values indicating greater compositional similarity between two samples.

* For each of the 17 participants in the control-first group, we compared the baseline sample to the end-of-control-phase sample, yielding 17 within-person similarity values

* Between-person similarity was calculated from all pairwise combinations of the 17 baseline samples across different participants, yielding $\binom{17}{2} = 136$ between-person similarity values. Summary statistics for both groups are shown below

```{r bc_similarity_prep}

# Read previous MAC data --------------------------------------------------

# Needed to extract treatment sequence
mac0 <- read_spss("../data/MAC Endpoint data with baseline values 102121.sav") %>% 
  janitor::clean_names()

# Variables needed
seq <- mac0 %>% 
  select(id, group, phase, visit, treatment)

# Distinct data with ID, Group
# n = 35
seq_distinct <- seq %>% 
  select(id, group) %>% 
  distinct()

# Use only control-first group
# Need to compare at baseline and at the end of control phase
ctrl_first_ids <- seq_distinct %>% filter(group == 2) %>% pull(id)
need_columns <- sprintf("%02.0f", ctrl_first_ids) %>% paste0("MAC", .)

# Read microbiome data ----------------------------------------------------

# 4166 rows x 112 columns
mb <- read_excel("../data/human_mac_study_abundance_data.xls")  

# Create abundance matrix
# Need rows = samples, columns = taxa
abundance_matrix <- mb %>% 
  select(starts_with(need_columns)) %>% 
  select(ends_with(c("v0", "v4"))) %>% 
  as.matrix() %>% 
  t()

```

```{r bc_similarity}

# Bray-Curtis dissimilarity / similarity -----------------------------------

# Dissimilarity score
bc_dissim <- vegdist(abundance_matrix, method = "bray") %>% 
  as.matrix()

# Similarity score
bc_sim <- 1 - bc_dissim

# Within-person BC similarity
sample_meta <- tibble(sample_id = rownames(bc_sim)) %>% 
  extract(sample_id, into = c("id_chr", "visit"),
          regex = "^(MAC\\d+)_p\\d+(v\\d+)$", remove = FALSE)

within_person_bc <- bc_sim %>% 
  as_tibble(rownames = "sample_1") %>% 
  pivot_longer(-sample_1, names_to = "sample_2", values_to = "bc_sim") %>% 
  left_join(sample_meta, by = c("sample_1" = "sample_id")) %>% 
  rename(id_1 = id_chr, visit_1 = visit) %>% 
  left_join(sample_meta, by = c("sample_2" = "sample_id")) %>% 
  rename(id_2 = id_chr, visit_2 = visit) %>% 
  filter(id_1 == id_2, visit_1 == "v0", visit_2 == "v4") %>% 
  select(id = id_1, bc_sim)

# Between-person BC similarity
between_person_bc <- bc_sim %>% 
  as_tibble(rownames = "sample_1") %>% 
  pivot_longer(-sample_1, names_to = "sample_2", values_to = "bc_sim") %>% 
  left_join(sample_meta, by = c("sample_1" = "sample_id")) %>% 
  rename(id_1 = id_chr, visit_1 = visit) %>% 
  left_join(sample_meta, by = c("sample_2" = "sample_id")) %>% 
  rename(id_2 = id_chr, visit_2 = visit) %>% 
  filter(visit_1 == "v0", visit_2 == "v0", id_1 < id_2) %>% 
  select(id_1, id_2, bc_sim)

# Summary statictics
bc_summary <- bind_rows(
  within_person_bc  %>% mutate(group = "Within-person"),
  between_person_bc %>% mutate(group = "Between-person")
) %>%
  group_by(group) %>%
  summarise(
    n      = n(),
    mean   = mean(bc_sim),
    sd     = sd(bc_sim),
    median = median(bc_sim),
    q1     = quantile(bc_sim, 0.25),
    q3     = quantile(bc_sim, 0.75)
  )

bc_summary %>%
  mutate(
    group  = group,
    n      = n,
    mean   = round(mean, 3),
    sd     = round(sd, 3),
    median = round(median, 3),
    q1     = round(q1, 3),
    q3     = round(q3, 3)
  ) %>%
  rename(
    Group    = group,
    N        = n,
    Mean     = mean,
    SD       = sd,
    Median   = median,
    Q1       = q1,
    Q3       = q3
  ) %>%
  knitr::kable(
    align   = "lcccccc"
  )

```

* Within-person similarity (mean = 0.655, median = 0.679) was substantially higher than between-person similarity (mean = 0.169, median = 0.157)
  * indicating that an individual's gut microbial composition remained far more similar to itself over time than to another participant's composition at a single timepoint

* This difference was confirmed formally using permutation MANOVA (PERMANOVA with 9999 permutations), testing whether subject ID explained variation in pairwise Bray-Curtis dissimilarity
  * Subject effect accounted for 88.4% of the total variance ($R^2 = 0.884$, $p < 0.0001$), providing strong statistical support for within-person compositional stability substantially exceeding between-person variability in this cohort

```{r bc_similarity_plot, fig.height = 4, fig.width = 5}

# PERMANOVA ---------------------------------------------------------------

# PERMANOVA based on dissimilarity
# Covariate table aligned to the rows of bc_dissim, in the same order
sample_info <- sample_meta %>% 
  filter(sample_id %in% rownames(bc_dissim)) %>% 
  arrange(match(sample_id, rownames(bc_dissim)))

# Does composition cluster by subject?
set.seed(123)
adonis_result <- adonis2(
  as.dist(bc_dissim) ~ id_chr,
  data = sample_info,
  permutations = 9999
)

# Boxplot with beeswarm ---------------------------------------------------

r2_val <- adonis_result$R2[1]
f_val  <- adonis_result$F[1]
p_val  <- adonis_result$`Pr(>F)`[1]

p_label <- if (p_val <= 0.0001) "p < 0.0001" else paste0("p = ", format(p_val, digits = 2))
perm_label <- sprintf("PERMANOVA: R² = %.3f, F = %.2f, %s", r2_val, f_val, p_label)

# Box plot
bind_rows(
  within_person_bc  %>% mutate(comparison = "within-person")  %>% select(comparison, bc_sim),
  between_person_bc %>% mutate(comparison = "between-person") %>% select(comparison, bc_sim)
) %>% 
  ggplot(aes(x = comparison, y = bc_sim)) +
  geom_boxplot() +
  geom_beeswarm(aes(color = comparison)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = NULL, 
    y = "Bray-Curtis Similarity",     
    title = "Bray-Curtis Similarity: Within- vs. Between-Person",
    subtitle = perm_label
  ) +
  theme_bw() +
  theme(legend.position = "none")

```

### Roseburia zero-inflation

* There are 11 of 35 participants (31.4%) with Roseburia_2 = 0 in both phases, and 6 more participants (17.1%) with Roseburia_2 = 0 in only one of the two phases -- bringing the total to 17 of 35 participants (49%) with at least one zero value
